State of the Practice for Lattice Boltzmann Method Software

### 2022-04-25 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Revisions completed for Section 2

We apply our methodology for evaluating the state of the practice of research
software \citep{SmithEtAl2021, SmithAndMichalski2022} to the specific domain of
LBM software.  Since we group best practices around conventional software
qualities, the first section below provides definitions for the qualities of
interest (Section~\ref{softwarequalities}), along with information on how to
measure these qualities. The following section
(Section~\ref{Sec_OverallProcess}) provides an overview of the steps we used to
select, measure and compare LBM software.  The remaining sections
(Sections~\ref{identifysoftware} to~\ref{SecDomainAnalysis}) provide details on
each step in the methodology.

@@

\begin{enumerate}
	\item List candidate software packages for the domain.
	(Section~\ref{identifysoftware})
	\item Filter the software package list. (Section~\ref{filtersoftware})
	\item Gather the source code and documentation for each software package.
	\item Collect quantitative measures from the project repositories.
	(Section~\ref{empiricalmeasures})
	\item Measure using the measurement template.  The full measurement template
	can be found in \citet{SmithEtAl2021}. (Section~\ref{empiricalmeasures})
	\item Use AHP to rank the software packages. (Section~\ref{AHP})
	\item Interview the developers. (Section~\ref{SecSurvey})
	\item Domain analysis. (Section~\ref{SecDomainAnalysis})
	\item Analyze the results and answer the research questions.
\end{enumerate}

@@

To answer~\rqref{RQ_WhatProjects} we identify existing LBM projects. We find the
candidate software through search engine queries targeting authoritative lists.
Scholarly articles, \href{https://github.com/} {GitHub} and
\href{https://swmath.org/} {swMATH} provide candidate LBM software. The Domain
Expert (Section~\ref{sec_vet_software_list}) helps with selecting the candidate
software.

We consider the following properties when creating the list and reviewing the
candidate software:

@@

The initial list had 45 packages, including some that we later found we could
not use because of incompleteness or an absence of publicly available source
code.

@@

To reduce the length of the list of software packages and to
answer~\rqref{RQ_WhatProjects}, we filter the initial list of 45 packages. We
apply the following filters in the priority order listed.

@@

For the third item in the above filter, we categorized software packages as
`alive' if documentation updates happened within the last 18 months, and `dead'
otherwise.

@@

Filtering by scope, usage, and age decreased the size of the list to 23
packages. We removed 22 packages for the following reasons: we cannot install
them, they are incomplete, source code is not publicly available, we are not
licensed to use them, or the project is out of scope. \citet{Michalski2021}
lists the eliminated software packages. Of the remaining 23 packages, we kept
some despite their `dead` status because of their presence on authoritative LBM
software lists and because of their surface excellence. After the initial
measurement exercise, which took place in mid-2020, we added one more package in
January 2022: Musubi. We found Musubi after our initial search, but we felt
obliged to include it since it clearly fits our criteria. Time constraints meant
we didn't update the previous manual data collection, but we did redo the
automated data collection.

The final list of 24 software packages is in two tables.
Table~\ref{alivepackages} lists packages that fell into the `alive' category as
of mid-2020 (January 2022 for Musubi), and Table~\ref{deadpackages} lists
packages that were `dead'. The tables include a hyperlink to each project and,
when available, citations for relevant publications.  The final list of software
packages shows considerable variation in purpose, size, user interfaces, and
programming languages. For example, the OpenLB software package is predominantly
a \CC{} package that uses hybrid parallelization and addresses many CFD problems
\citep{heuveline2009towards}. The software package pyLBM is an all-in-one Python
language package for numerical simulations \citep{graille2017pylbm}. ESPResSo
consists of extensible \CC{} and Python code specifically for research on soft
matter \citep{weik2019espresso}. The HemeLB package simulates fluid flow in
several medical domains and is written predominantly in C, \CC, and Python
\citep{mazzeo2008hemelb}.

@@

We rank the projects by how well they follow best practices
(\rqref{RQ_HighestQuality}) via a measurement template, as described in
\citet{SmithEtAl2021}.  For each software package (each column in the template),
we fill in the rows of the template. This process takes about two hours per
package, with a cap of four hours. The time constraint is necessary so that the
workload is feasible for a team as small as one, given our aim to cap the
measurement phase at 160 person-hours \citep{SmithEtAl2021}.  An excerpt of the
template, in spreadsheet form, is shown in
Figure~\ref{measurement_template_image}.

@@

We designed the questions to be unambiguous, quantifiable, and measurable with
limited time and domain knowledge. We categorize them by quality, with an extra
category for summary information (Figure~\ref{measurement_template_image}). The
summary section provides general data, such as the software name, number of
developers, etc. \citet{gewaltig2012quality} provide us with our software
categories. Public means software available to all, while private means software
available only to a specific group.  The concept category defines software
written to demonstrate algorithms or concepts. The three categories of
development models are open source, where source code is freely available under
an open-source license; free-ware, which freely provides a binary or executable,
but not source code; and commercial, where the user must pay for the software.

Several of the qualities use the word ``surface'' to highlight that, for these
qualities, the best that we can do is a shallow measure. For instance, we do not
conduct experiments to measure usability directly. Instead, we look for
usability considerations by searching for cues in the documentation, such as the
presence of getting started instructions, a user manual or documentation of
expected user characteristics.

We used tools to automatically measure the number of files, number of lines of
code (LOC), percentage of closed issues, etc. GitStats measures each package's
GitHub repository for the number of binary files, the number of added and
deleted lines, and the number of commits over varying time intervals. Sloc Cloc
and Code (scc) measures the number of text files and the number of total, code,
comment, and blank lines in each GitHub repository.

As in \citet{SmithEtAl2016}, we used Virtual machines (VMs) to provide an
optimal testing environment for each package. We used VMs to create fresh
environments, eliminating worry about existing libraries and conflicts.
Moreover, when the tests are complete, the VM can be deleted, without impacting
the host operating system. The most significant advantage of using VMs is to
level the playing field. Every software install starts from a clean slate, which
removes ``works-on-my-computer'' errors.

@@

Our ranking uses a tool we developed for conducting the AHP analysis. The tool
includes a sensitivity analysis to ensure that the software package rankings are
appropriate with respect to the uncertainty of the quality scores. For the
sensitivity analysis, we modified the score by 10\% for each package and
verified that the overall ranking was stable.  The
\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2020/LBM/README.txt}{README}
file for the tool includes requirements and usage information.

@@

Several of the research question (\rqref{RQ_CompareToolsProjMngmnt},
\rqref{RQ_CompareMethodologies}, \rqref{RQ_PainPoints}, \rqref{RQ_Concerns} and
\rqref{RQ_Recommend}) require going beyond the quantitative data from the
measurement template. To gain insight, we interviewed developers using a list of
20 questions from \citet{SmithEtAl2021}. The questions cover the background of
the development teams, the interviewees, and the software itself. We ask the
developers how they organize their projects. We also ask about their
understanding of the users. Some questions focus on the current and past
difficulties, and the solutions the team has found or will try. We also discuss
the importance of, and the current situation for, documentation. A few questions
focus on specific software qualities, such as maintainability,
understandability, usability, and reproducibility. The interviews are
semi-structured based on the question list; we ask follow-up questions when
necessary. Based on our experience, the interviewees often raise exciting and
unexpected ideas.

@@

The domain analysis (also called a commonality analysis) consists of
understanding the commonalities, variabilities, and common terminology for a
program family \citep{weiss1997defining}. Commonalities are goals, theories,
models, definitions, and assumptions that are common between family members.
Variabilities are goals, theories, models, definitions, and assumptions that
differ between family members. Associated with each variability are parameters
of variation, which summarize the possible values for that variability, along
with their potential binding time. The binding time is when we specify a value
for a variability. Binding time options include specification time, build time
(when the program compiles) and run time (when the code executes). For the
current assessment, the packages all have the commonality of using LBM
techniques to simulate the motion of fluids. Variabilities for LBM packages
include the following: the dimension of problems they simulate, whether (and
how) they support parallel computation, whether compressibility can be
considered, whether reflexive boundaries can be employed, whether the simulation
handles multi-fluids, whether the simulation handles turbulence, and whether
complex geometries can be simulated.

To keep the current presentation focused, we only present the results of our
shallow domain analysis. We distinguish programs by their variabilities, which
for research software are often assumptions. Table~\ref{tbl_features} lists the
variabilities as columns, with the software alphabetically ordered.
\citet{Michalski2021} provides a detailed analysis of the commonalities and
variabilities between LBM packages.

@@

The Domain Expert has a central role in verifying the list of LBM packages. In
advance of the first meeting with the Domain Expert, we asked them to create a
list of top software packages in the domain.  We did this to help the expert get
in the right mindset for our meeting.  Moreover, by doing the exercise in
advance, we avoid the potential pitfall of the expert approving the discovered
list of software without giving it adequate thought. We asked the Domain Expert,
via virtual meetings, to vet the collected data and analysis. In particular, we
asked them to vet the proposed list of software packages and the AHP ranking.

>>>>

We apply our methodology for evaluating the state of the practice of research
software \citep{SmithEtAl2021, SmithAndMichalski2022} to the specific domain of
LBM software.  Since we group best practices around conventional software
qualities, the first section below provides definitions for the qualities of
interest (Section~\ref{softwarequalities}), along with information on how to
measure these qualities. The following section
(Section~\ref{Sec_OverallProcess}) provides an overview of the steps we used to
select, measure and compare LBM software.  The remaining sections
(Sections~\ref{identifysoftware} to~\ref{sec_vet_software_list}) provide details
on each step in the methodology.

@@

\begin{enumerate}
	\item List candidate software packages for the domain.
	(Section~\ref{identifysoftware})
	\item Filter the software package list. (Section~\ref{filtersoftware})
	\item Gather the source code and documentation for each software package.
	\item Collect quantitative measures from the project repositories.
	(Section~\ref{empiricalmeasures})
	\item Measure using the measurement template.  The full measurement template
	can be found in \citet{SmithEtAl2021}. (Section~\ref{empiricalmeasures})
	\item Rank the software using AHP. (Section~\ref{AHP})
	\item Interview the developers. (Section~\ref{SecSurvey})
	\item Analyze the domain. (Section~\ref{SecDomainAnalysis})
	\item Analyze the results and answer the research questions.
\end{enumerate}

@@

To answer~\rqref{RQ_WhatProjects} we identify existing LBM projects through
search engine queries targeting authoritative lists. We find the software in
scholarly articles, \href{https://github.com/} {GitHub} and
\href{https://swmath.org/} {swMATH}. The Domain Expert
(Section~\ref{sec_vet_software_list}) also helps with selecting the candidate
software.  We build our list based on the following criteria:

@@

\noindent Our initial list had 45 packages, including some that we later found
to be unusable because of incompleteness, or an absence of publicly available
source code.

@@

To reduce the length of the list and to answer~\rqref{RQ_WhatProjects}, we
filter the initial list of 45 packages. We apply the following filters in the
priority order listed.

@@

For the third item in the above filter, we categorize software packages as
`alive' if documentation updates happened within the last 18 months, and `dead'
otherwise.

@@

Filtering by scope, usage, and age decreased the size of the list to 23
packages. We removed 22 packages for the following reasons: we cannot install
them, they are incomplete, source code is not publicly available, we are not
licensed to use them, or the project is out of scope. \citet{Michalski2021}
lists the eliminated software packages. Of the remaining 23 packages, we kept
some despite their `dead` status because of their presence on authoritative LBM
software lists and because of their surface excellence. After the initial
measurement exercise, which took place in mid-2020, we added one more package in
January 2022: Musubi. We found Musubi after our initial search, but we felt
obliged to include it since it clearly fits our criteria. Time constraints meant
we could not update the previous manual data collection, but we did redo the
automated data collection.

The final list of 24 software packages is in two tables.
Table~\ref{alivepackages} lists packages that fall into the `alive' category as
of mid-2020 (January 2022 for Musubi), and Table~\ref{deadpackages} lists
packages that are `dead'. The tables include a hyperlink to each project and,
when available, citations for relevant publications.  The final list of software
packages shows considerable variation in purpose, size, user interfaces, and
programming languages. For example, the OpenLB software package is predominantly
a \CC{} package that uses hybrid parallelization and addresses many CFD problems
\citep{heuveline2009towards}. The software package pyLBM is an all-in-one Python
language package for numerical simulations \citep{graille2017pylbm}. ESPResSo
consists of extensible \CC{} and Python code specifically for research on soft
matter \citep{weik2019espresso}. The HemeLB package simulates fluid flow in
several medical domains and is written predominantly in C, \CC, and Python
\citep{mazzeo2008hemelb}.

@@

We use a measurement template to rank the projects by how well they follow best
practices (\rqref{RQ_HighestQuality}), as described in \citet{SmithEtAl2021}.
For each software package (each column in the template), we fill in the rows of
the template. This process takes about two hours per package, with a cap of four
hours. The time constraint is necessary so that the workload is feasible for a
team as small as one, given our aim to cap the measurement phase at 160
person-hours \citep{SmithEtAl2021}.  An excerpt of the template, in spreadsheet
form, is shown in Figure~\ref{measurement_template_image}.

@@

We designed the questions to be unambiguous, quantifiable, and measurable with
limited time and domain knowledge. We categorize them by quality, with an extra
category for summary information (Figure~\ref{measurement_template_image}). The
summary section provides general data, such as the software name, number of
developers, etc. \citet{gewaltig2012quality} provide us with our software
categories. Public means software available to all, while private means software
available only to a specific group.  The concept category defines software
written to demonstrate algorithms or concepts. The three categories of
development models are open source, where source code is freely available under
an open-source license; free-ware, which freely provides a binary or executable,
but not source code; and commercial, where the user must pay.

Several of the qualities use the word ``surface'' to highlight that, for these
qualities, the best that we can do is a shallow measure. For instance, we do not
conduct experiments to measure usability directly. Instead, we look for
usability considerations by searching for cues in the documentation, such as the
presence of getting started instructions, a user manual, or documentation of
expected user characteristics.

We used tools to automatically measure the number of files, number of lines of
code (LOC), percentage of closed issues, etc.
\href{https://github.com/tomgi/git_stats}{GitStats} \citep{Gieniusz2019}
measures each package's GitHub repository for the number of binary files, the
number of added and deleted lines, and the number of commits over varying time
intervals. \href{https://github.com/boyter/scc}{Sloc Cloc and Code (scc)}
\citep{Boyter2021} measures the number of text files and the number of total,
code, comment, and blank lines in each GitHub repository.

As in \citet{SmithEtAl2016}, we used Virtual Machines (VMs) to provide an
optimal testing environment for each package. We used VMs to create fresh
environments, eliminating any worries about existing libraries and conflicts.
Moreover, when the tests are complete, the VM can be deleted, without impacting
the host operating system. The most significant advantage of using VMs is that
they level the playing field. Every software install starts from a clean slate,
which removes ``works-on-my-computer'' errors.

@@

We use our own simple tool for conducting the AHP analysis. The tool includes a
sensitivity analysis to ensure that the software package rankings are
appropriate with respect to the uncertainty of the quality scores. For the
sensitivity analysis, we modified the score by 10\% for each package and
verified that the overall ranking was stable.  The
\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2020/LBM/README.txt}{README}
file for the tool includes requirements and usage information.

@@

Two of the research question (\rqref{RQ_PainPoints} and \rqref{RQ_Concerns})
require going beyond the quantitative data from the measurement template. To
gain insight, we interview developers using a list of 20 questions from
\citet{SmithEtAl2021}. The questions cover the background of the development
teams, the interviewees, and the software itself. We ask the developers how they
organize their projects and about their understanding of the users. Some
questions focus on the current and past difficulties, and the solutions the team
has found or will try. We also discuss the importance of, and the current
situation for, documentation. A few questions focus on specific software
qualities, such as maintainability, understandability, usability, and
reproducibility. The interviews are semi-structured based on the question list;
we ask follow-up questions when necessary. Based on our experience, the
interviewees often raise exciting and unexpected ideas.

@@

The domain analysis (also called a commonality analysis) consists of
understanding the commonalities, variabilities, and common terminology for a
program family \citep{weiss1997defining}. Commonalities are goals, theories,
models, definitions, and assumptions that are common between family members.
Variabilities are goals, theories, models, definitions, and assumptions that
differ between family members. Associated with each variability are parameters
of variation, which summarize the possible values for that variability, along
with their potential binding time. The binding time is when a variability is
assigned a value. Binding time options include specification time, build time
(when the program compiles) and run time (when the code executes). For the
current assessment, the packages all have the commonality of using LBM
techniques to simulate the motion of fluids. Variabilities for LBM packages
include the following: the dimension of problems they simulate, whether (and
how) they support parallel computation, whether compressibility can be
considered, whether reflexive boundaries can be employed, whether the simulation
handles multi-fluids, whether the simulation handles turbulence, and whether
complex geometries can be simulated.

To focus the current presentation, we only present the results of our shallow
domain analysis. We distinguish programs by their variabilities, which for
research software are often assumptions. Table~\ref{tbl_features}, which is
sorted alphabetically, lists the variabilities as columns. \citet{Michalski2021}
provides a more detailed analysis of the commonalities and variabilities between
LBM packages.

@@

The Domain Expert has a central role in verifying the list of packages. In
advance of the first meeting with the Domain Expert, we ask them to create a
list of top software packages in the domain.  We do this to help the expert get
in the right mindset for our meeting.  Moreover, by doing the exercise in
advance, we avoid the potential pitfall of the expert approving the discovered
list of software without giving it adequate thought. We ask the Domain Expert to
vet the collected data and analysis. In particular, we ask them to vet the
proposed list of software packages and the AHP ranking.
